---
title: "Managing and Understanding Data"
author: "Escribir vuestro nombre y apellidos"
date: '`r format(Sys.Date(),"%e de %B, %Y")`'
---

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)

```

```{r libraries, include=FALSE}
# Install packages
# Load packages
# ...

library(knitr)
```


```{r input, include=FALSE}
# Input / Output variables
# Tuning parameters
# ...
file1 <- "usedcars.csv"

```




> * Primera toma de contacto con un informe dinámico donde se muestra algunas de sus caracteristicas.
> * Son diferentes trozos de un libro.
> * Se lee un archivo csv


-----------------------------------------------------------

`r Sys.Date()`

By the end of this notes, you will understand:

* The basic R data structures and how to use them to store and extract data

* How to get data into R from a variety of source formats
* Common methods for understanding and visualizing complex data


# R data structures

The R data structures used most frequently in machine learning are *vectors*, *factors*, *lists*, *arrays*,
and *data frames*.


To find out more about machine learning see [@andrieu2003introduction;@goldberg1988genetic].

## Vectors 

The fundamental R data structure is the **vector**, which stores an ordered set of
values called **elements**. A vector can contain any number of elements. However, all
the elements must be of the same type; for instance, a vector cannot contain both
numbers and text.

There are several vector types commonly used in machine learning:`integer`(numbers without decimals), `numeric` (numbers with decimals), `character` (text data),
or `logical` (`TRUE` or `FALSE` values). There are also two special values: `NULL`, which is
used to indicate the absence of any value, and `NA`, which indicates a missing value.


...

...

...


Create vectors of data for three medical patients:

```{r chunck1}
# create vectors of data for three medical patients
subject_name <- c("John Doe", "Jane Doe", "Steve Graves")
temperature <- c(98.1, 98.6, 101.4)
flu_status <- c(FALSE, FALSE, TRUE)
```
